---
title: "Mating Short Communication Stats"
author: Michelle Zoë Hotchkiss
output: html_notebook
editor_options: 
  chunk_output_type: console
---

# Set Up

Load packages

```{r}
library(tidyverse)
library(MetBrewer)
library(car)
library(lmtest)
library(PerformanceAnalytics)
```


Load mating and survival data data.

```{r}
matingdata <- read.csv("data/matingdata_allqueens.csv", stringsAsFactors = TRUE, na.strings=c(""))

str(matingdata)

matingdata$emergence_weight <- as.numeric(matingdata$emergence_weight)

matingdata$end_weight <- as.numeric(matingdata$end_weight)

matingdata$hib_weightloss <- as.numeric(matingdata$hib_weightloss)

matingdata$rec_weightdiff <- as.numeric(matingdata$rec_weightdiff)

```


# Figure 1: Boxplot of mating type vs relative weight difference

```{r}
md_summary <- matingdata %>% 
  filter(!(is.na(mating_type))) %>% 
  group_by(mating_type) %>% 
  tally()

md_summary
```

Boxplot:

```{r}
matingdata %>% 
  filter(!(is.na(mating_type))) %>% 
  ggplot(aes(x=mating_type, y=rel_prehib_weightdiff, fill=mating_type)) + 
  geom_boxplot(alpha=0.8)+
  geom_jitter(shape=16, position=position_jitter(0), size=2)+
  #geom_text(data = md_summary, aes(mating_type, Inf, label = n), vjust = 2)+
  theme_classic()+
  theme(text = element_text(size = 16), axis.text.x = element_text(size = 16, color = "black"),
        axis.text.y = element_text(size = 16, color = "black"),
        legend.position = "none",
        axis.title.y = element_text(size = 16, margin = margin(t = 0, r = 20, b = 0, l = 0)),
        axis.title.x = element_text(size = 16, margin = margin(t = 20, r = 0, b = 0, l = 0)))+
  scale_x_discrete(name="Mating Status")+
  scale_y_continuous(name ="Relative Difference in Weight \nFrom 0 to 10 Days Post-Eclosion")+
  scale_fill_manual(values = met.brewer("Homer1",2))

#Homer1
```


#Figure of relative weight difference vs. day 0 weight - this is more true to the model, but a little more complicated. Do you prefer this or the boxplot?

```{r}
matingdata %>% 
  ggplot(aes(x=day_0_weight, y=rel_prehib_weightdiff, color=mating_type)) + 
  geom_point(alpha=0.8)+
  geom_smooth()+
  theme_classic()+
  theme(text = element_text(size = 16), axis.text.x = element_text(size = 16, color = "black"),
        axis.text.y = element_text(size = 16, color = "black"),
        legend.position = "none",
        axis.title.y = element_text(size = 16, margin = margin(t = 0, r = 20, b = 0, l = 0)),
        axis.title.x = element_text(size = 16, margin = margin(t = 20, r = 0, b = 0, l = 0)))+
  scale_x_continuous(name="Day 0 Weight (mg)")+
  scale_y_continuous(name ="Relative Difference in Weight \nFrom 0 to 10 Days Post-Eclosion")
```


#Figure of day 0 vs. day 10 weight

```{r}
matingdata %>% 
  ggplot(aes(x=day_0_weight, y=day_10_weight, color=mating_type)) + 
  geom_point(alpha=0.8)+
  geom_smooth()+
  theme_classic()+
  theme(text = element_text(size = 16), axis.text.x = element_text(size = 16, color = "black"),
        axis.text.y = element_text(size = 16, color = "black"),
        legend.position = "none",
        axis.title.y = element_text(size = 16, margin = margin(t = 0, r = 20, b = 0, l = 0)),
        axis.title.x = element_text(size = 16, margin = margin(t = 20, r = 0, b = 0, l = 0)))+
  scale_x_continuous(name="Day 0 Weight (mg)")+
  scale_y_continuous(name ="Day 10 Weight (mg)")
```

# Weight loss over hibernation


```{r}
matingdata %>% 
  filter(!(is.na(mating_type))) %>% 
  ggplot(aes(x=mating_type, y=hib_weightloss, fill=mating_type)) + 
  geom_boxplot(alpha=0.8)+
  geom_jitter(shape=16, position=position_jitter(0))+
  theme_classic()+
  theme(text = element_text(size = 16), axis.text.x = element_text(size = 16, color = "black"),
        axis.text.y = element_text(size = 16, color = "black"),
        legend.position = "none",
        axis.title.y = element_text(size = 16, margin = margin(t = 0, r = 20, b = 0, l = 0)),
        axis.title.x = element_text(size = 16, margin = margin(t = 20, r = 0, b = 0, l = 0)))+
  scale_x_discrete(name="Mating Type")+
  scale_y_continuous(name ="Hibernation Weight Loss")+
  scale_fill_manual(values = met.brewer("Homer1",2))

#Homer1
```

Line
```{r}
matingdata %>% 
  ggplot(aes(x=day_0_weight, y=hib_weightloss, color=mating_type)) + 
  geom_point(alpha=0.8)+
  geom_smooth()+
  theme_classic()+
  theme(text = element_text(size = 16), axis.text.x = element_text(size = 16, color = "black"),
        axis.text.y = element_text(size = 16, color = "black"),
        legend.position = "none",
        axis.title.y = element_text(size = 16, margin = margin(t = 0, r = 20, b = 0, l = 0)),
        axis.title.x = element_text(size = 16, margin = margin(t = 20, r = 0, b = 0, l = 0)))+
  scale_x_continuous(name="Day 0 Weight (mg)")+
  scale_y_continuous(name ="Hibernation Weight Loss")
```



# Statistical Analyses

Means

```{r}
matingdata %>% 
  filter(!(is.na(mating_type))) %>% 
  group_by(mating_type) %>% 
  summarise(mean(rel_prehib_weightdiff))


matingdata %>% 
  filter(!(is.na(mating_type))) %>% 
  group_by(mating_type) %>% 
  summarise(median(rel_prehib_weightdiff))
```


## ANCOVA 

### Set initial model

```{r}
matingmodel1 <- lm(rel_prehib_weightdiff ~ mating_type + day_0_weight, data=matingdata)

#for this step, i predict that mated bees will gain more weight regardless of how much they weigh at day 0, so i did not include an interaction. also i don't think we have the power to include an interaction (i.e., so few low weight mated and so few mated in general), so i think it's better to fit the simpler model. perhaps more balanced designs later can detect an interaction
```

###Test assumptions

Visually:

```{r}
opar <- par(mfrow = c(2, 2))
plot(matingmodel1)
par(opar)
```

With statistical tests:

```{r}
#Test for normality
shapiro.test(residuals(matingmodel1))
#W = 0.94973, p-value = 0.0195

#Test for homoscedasticity
bptest(matingmodel1)
#BP = 3.6985, df = 2, p-value = 0.5683
```

Assumptions are not completely met, normality is violated.

### Examine results of model 1

```{r}
car::Anova(matingmodel1, type = 3)

#Anova Table (Type III tests)

#Response: rel_prehib_weightdiff
#             Sum Sq Df F value    Pr(>F)    
#(Intercept)  0.06742  1 10.2507  0.002377 ** 
#mating_type  0.11879  1 18.0596 9.326e-05 ***
#day_0_weight 0.00358  1  0.5447  0.463953    
#Residuals    0.32887 50                      
#---
#Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
```


## Relationship between day mated and weight gain

```{r}
matingdata %>% 
  drop_na(mating_type) %>% 
  ggplot(aes(x=day_mated, y=rel_prehib_weightdiff, color=mating_type)) + 
  geom_point()
```

Doesn't look like much, so i think we're in the clear, but we can run stats on it later to be sure. Is this a question that you had while reading?


## Mortality Analysis

### Figure

Figure of percent mortality by mating at 4 months

```{r}
matingdata %>% 
  filter(treatment == "Recovery C"|treatment=="Recovery G"| treatment == "4 month") %>% 
  drop_na(mating_type) %>% 
  group_by(status, mating_type) %>% 
  summarise(n=n()) %>% 
  ggplot(aes(x=status, y=n, fill=status)) + 
  geom_bar(stat='identity')+
  facet_grid(. ~ mating_type) +
  theme_classic()+
  theme(text = element_text(size = 16), axis.text.x = element_text(size = 16),
        legend.position = "none",
        axis.title.y = element_text(size = 16, margin = margin(t = 0, r = 20, b = 0, l = 0)),
        axis.title.x = element_text(size = 16, margin = margin(t = 20, r = 0, b = 0, l = 0)))+
  scale_x_discrete(name="Status")+
  scale_y_continuous(name ="Number of Bees", limits =c(0,20), expand = c(0, 0))+
  scale_fill_manual(values = met.brewer("Homer1",2))
```

Point graph of essentially the same info, I think I like this one better because you can see the gap and difference between light and heavy queens

```{r}
matingdata %>% 
  filter(!(is.na(status))) %>% 
  ggplot(aes(x=day_0_weight, y=status, color=mating_type)) + 
  geom_point(position = position_jitter(w = 0, h = 0.05, seed = 22),size = 3)+
  theme_classic()+
  theme(text = element_text(size = 16), 
        axis.text.x = element_text(size = 16, color = "black"),
        axis.text.y = element_text(color = "black"),
        legend.position = "right",
        axis.title.y = element_text(size = 16, margin = margin(t = 0, r = 20, b = 0, l = 0)),
        axis.title.x = element_text(size = 16, margin = margin(t = 20, r = 0, b = 0, l = 0)))+
  scale_x_continuous(name="Day 0 Weight (mg)")+
  scale_y_discrete(name ="Survival Status", limits = rev)+
  scale_color_manual(name = "Mating Status", values = met.brewer("Homer1",2))
```


With all end points

```{r}
matingdata %>% 
  drop_na(mating_type) %>% 
  drop_na(status) %>% 
  group_by(status, mating_type) %>% 
  summarise(n=n()) %>% 
  ggplot(aes(x=status, y=n, fill=status)) + 
  geom_bar(stat='identity')+
  facet_grid(. ~ mating_type) +
  theme_classic()+
  theme(text = element_text(size = 16), axis.text.x = element_text(size = 16),
        legend.position = "none",
        axis.title.y = element_text(size = 16, margin = margin(t = 0, r = 20, b = 0, l = 0)),
        axis.title.x = element_text(size = 16, margin = margin(t = 20, r = 0, b = 0, l = 0)))+
  scale_x_discrete(name="Status")+
  scale_y_continuous(name ="Number of Bees", limits =c(0,30), expand = c(0, 0))+
  scale_fill_manual(values = met.brewer("Homer1",2))
```

### Survival Analysis with Multiple Logistic Regression

Using code from https://rcompanion.org/rcompanion/e_07.html, assumptions from http://www.biostathandbook.com/multiplelogistic.html

Our variables have to be made numeric. Start by selecting just the variables that we want.

```{r}
matingdata_survival_numeric <- matingdata %>% 
  select(day_0_weight, day_10_weight, mating_type, prehib_weightdiff, status)

matingdata_survival_numeric$prehib_weightdiff <- as.numeric(matingdata_survival_numeric$prehib_weightdiff)

str(matingdata_survival_numeric)
```

Then make everything numeric

```{r}
matingdata_survival_numeric$mating_type_num <- as.numeric(matingdata_survival_numeric$mating_type)
#1 = mated, 2 = unmated

matingdata_survival_numeric$status_num <- as.numeric(matingdata_survival_numeric$status)
#1 = alive, 2 = dead

#change the 2 for dead to a zero
matingdata_survival_numeric$status_num <-
  replace(matingdata_survival_numeric$status_num,matingdata_survival_numeric$status_num == 2,0)

#use str to double check that variables are coded properly
str(matingdata_survival_numeric)
```

Make variable for relative pre-hibernation weight difference

```{r}
matingdata_survival_numeric$rel_prehib_weightdiff <- ((matingdata_survival_numeric$day_10_weight-matingdata_survival_numeric$day_0_weight)/matingdata_survival_numeric$day_0_weight)
  
str(matingdata_survival_numeric)
```


####Correlation Analysis

Remove the factors for the correlation analysis.

```{r}
matingdata_survival_numeric_nofac <- matingdata_survival_numeric[-c(3,5)]
str(matingdata_survival_numeric_nofac)
```


Examine correlation among variables

**THIS IS THE STEP FOR CORRELATION TESTING**

```{r}
chart.Correlation(matingdata_survival_numeric_nofac, method = "spearman", histogram = TRUE)
```


#### Survival Model 1

```{r}
survival_model1 <- glm(status_num ~ mating_type_num + day_0_weight + rel_prehib_weightdiff, data = matingdata_survival_numeric_nofac, family= binomial)

summary(survival_model1)

#Deviance Residuals: 
#    Min       1Q   Median       3Q      Max  
#-2.1250  -0.7497   0.3423   0.6258   1.8828  

#Coefficients:
#                        Estimate Std. Error z value Pr(>|z|)   
#(Intercept)            2.820e+01  4.827e+03   0.006  0.99534   
#mating_type_num       -1.668e+01  2.414e+03  -0.007  0.99449   
#day_0_weight           8.884e-03  2.736e-03   3.247  0.00117 **
#rel_prehib_weightdiff  1.342e+01  5.997e+00   2.238  0.02522 * 
#---
#Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

#(Dispersion parameter for binomial family taken to be 1)

#    Null deviance: 70.252  on 52  degrees of freedom
#Residual deviance: 47.177  on 49  degrees of freedom
#AIC: 55.177

#Number of Fisher Scoring iterations: 17
```

Variance inflation factors to detect multi-colinearity

```{r}
vif(survival_model1)

#      mating_type_num          day_0_weight rel_prehib_weightdiff 
#             1.000000              1.133545              1.133545
```

No evidence of colinearity

Plot residuals

```{r}
plot(fitted(survival_model1),
     rstandard(survival_model1))
```

Check for overdispersion

Overdispersion is a situation where the residual deviance of the glm is large relative to the residual degrees of freedom. 
One guideline is that if the ratio of the residual deviance to the residual degrees of freedom exceeds 1.5, then the model is overdispersed. 

```{r}
summary(survival_model1)$deviance / summary(survival_model1)$df.residual

#0.962797
```

#### Survival Model 2 and 3 - does removed mating_type or rel_prehib make a difference for significance?

```{r}
survival_model2 <- glm(status_num ~ mating_type_num + day_0_weight, data = matingdata_survival_numeric_nofac, family= binomial)


summary(survival_model2)
```

```{r}
survival_model3 <- glm(status_num ~ day_0_weight + rel_prehib_weightdiff, data = matingdata_survival_numeric_nofac, family= binomial)


summary(survival_model3)
```



