---
title: "Mating Short Communication Stats"
author: Michelle Zoë Hotchkiss
output: html_notebook
editor_options: 
  chunk_output_type: console
---

# Set Up

Load packages.

```{r}
library(tidyverse)
library(ggpubr)
library(car)
library(lmtest)
library(PerformanceAnalytics)
library(MetBrewer)
```


Load data and recode variables.

```{r}
matingdata <- read.csv("data/matingdata.csv", stringsAsFactors = TRUE, na.strings=c(""))
matingdata <- matingdata[-c(54,55,56,57,58,59), ]

str(matingdata)

matingdata$emergence_weight <- as.numeric(matingdata$emergence_weight)

matingdata$end_weight <- as.numeric(matingdata$end_weight)

matingdata$hib_weightloss <- as.numeric(matingdata$hib_weightloss)
```


# Figure 1: Boxplot of mating type vs relative weight difference

```{r}
matingdata %>% 
  filter(!(is.na(mating_type))) %>% 
  ggplot(aes(x=mating_type, y=rel_prehib_weightdiff, fill=mating_type, color=mating_type)) + 
  geom_boxplot(alpha=0.4, lwd=0.7)+
  geom_jitter(shape=16, position=position_jitter(0.09), size=3)+
  #geom_text(data = md_summary, aes(mating_type, Inf, label = n), vjust = 2)+
  theme_classic()+
  theme(text = element_text(size = 16), axis.text.x = element_text(size = 16, color = "black"),
        axis.text.y = element_text(size = 16, color = "black"),
        legend.position = "none",
        axis.title.y = element_text(size = 16, margin = margin(t = 0, r = 20, b = 0, l = 0)),
        axis.title.x = element_text(size = 16, margin = margin(t = 20, r = 0, b = 0, l = 0)))+
  scale_x_discrete(name="Mating Status")+
  scale_y_continuous(name ="Relative Difference in Weight \nFrom 0 to 10 Days Post-Eclosion")+
  scale_fill_met_d("Homer1")+
  scale_color_met_d("Homer1")
```


#Figure 2a: Relative weight loos during hibernation vs. Weight at hibernation onset

```{r}
#make a variable for relative weight loss over hibernation

matingdata$rel_hib_weightloss <- (matingdata$day_10_weight-matingdata$emerg_weight)/matingdata$day_10_weight

#create plot

fig2a <- matingdata %>% 
  filter(!(is.na(status))) %>% 
  filter(!(is.na(rel_hib_weightloss))) %>% 
  ggplot(aes(x=day_10_weight, y=rel_hib_weightloss, color=mating_type)) + 
  geom_point(position = position_jitter(w = 0, h = 0.05, seed = 22),size = 4,shape = 17)+
  theme_classic()+
  theme(text = element_text(size = 16), 
        axis.text.x = element_text(size = 16, color = "black"),
        axis.text.y = element_text(color = "black"),
        legend.position = "none",
        axis.title.y = element_text(size = 16, margin = margin(t = 0, r = 20, b = 0, l = 0)),
        axis.title.x = element_text(size = 16, margin = margin(t = 20, r = 0, b = 0, l = 0)))+
  scale_x_continuous(name="Weight at Hibernation Onset (mg)")+
  scale_y_continuous(name ="Relative Weight Loss During Hibernation")+
  scale_color_manual(name = "Mating Status", values = met.brewer("Homer1",2))
```



#Figure 2b: Survival vs. weight at eclosion


```{r}
fig2b <- matingdata %>% 
  filter(!(is.na(status))) %>% 
  ggplot(aes(x=day_0_weight, y=status, color=mating_type, size = rel_prehib_weightdiff)) + 
  geom_point(position = position_jitter(w = 0, h = 0.15, seed = 22))+
  theme_classic()+
  theme(text = element_text(size = 16), 
        axis.text.x = element_text(size = 16, color = "black"),
        axis.text.y = element_text(color = "black"),
        legend.position = "right",
        legend.title=element_text(size=14),
        axis.title.y = element_text(size = 16, margin = margin(t = 0, r = 20, b = 0, l = 0)),
        axis.title.x = element_text(size = 16, margin = margin(t = 20, r = 0, b = 0, l = 0)))+
  guides(colour = guide_legend(override.aes = list(size=4)), 
         size = guide_legend(override.aes = list(colour = "#00b6e0")))+
  scale_x_continuous(name="Weight at Eclosion (mg)")+
  scale_y_discrete(name ="Survival Status", limits = rev)+
  scale_color_manual(name = "Mating Status", values = met.brewer("Homer1",2))+
  scale_size_continuous(name = "Relative \nPre-Hibernation \nWeight \nDifference")
```


#Assembling Figure 2

```{r}
ggarrange(fig2a, NULL, fig2b, 
          nrow=1, 
          widths = c(3,0.2,4), 
          labels = c("A","","B"),
          font.label = list(size = 20))
```


# Model 1: Relative pre-hibernation weight difference

Examine means.

```{r}
matingdata %>% 
  filter(!(is.na(mating_type))) %>% 
  group_by(mating_type) %>% 
  summarise(mean(rel_prehib_weightdiff))
```

## Define model

```{r}
matingmodel1 <- lm(rel_prehib_weightdiff ~ mating_type + day_0_weight, data=matingdata)
```

##Test assumptions

Visually:

```{r}
opar <- par(mfrow = c(2, 2))
plot(matingmodel1)
par(opar)
```

With statistical tests:

```{r}
#Test for normality
shapiro.test(residuals(matingmodel1))

#Test for homoscedasticity
bptest(matingmodel1)
```

Normality is violated in the formal test, but visually the data looks okay.

## Results of Model 1

```{r}
car::Anova(matingmodel1, type = 3)
summary(matingmodel1)
```


# Model 2: Relative weight loss during hibernation

If figure code was not run, make variable for relative weight loss during hibernation. 

```{r}
matingdata$rel_hib_weightloss <- (matingdata$day_10_weight-matingdata$emerg_weight)/matingdata$day_10_weight
```

Examine means.

```{r}
matingdata %>% 
  filter(!(is.na(mating_type))) %>% 
  filter(!(is.na(rel_hib_weightloss))) %>% 
  group_by(mating_type) %>% 
  summarise(mean(rel_hib_weightloss))
```

## Define model.

```{r}
matingmodel2 <- lm(rel_hib_weightloss ~ mating_type + day_10_weight, data=matingdata)
```

##Test assumptions

Visually:

```{r}
opar <- par(mfrow = c(2, 2))
plot(matingmodel2)
par(opar)
```

With statistical tests:

```{r}
#Test for normality
shapiro.test(residuals(matingmodel2))

#Test for homoscedasticity
bptest(matingmodel2)
```

Assumptions are met.

## Results of Model 2

```{r}
car::Anova(matingmodel2, type = 3)
summary(matingmodel2)
```


# Model 3: Survival Analysis

Using code from https://rcompanion.org/rcompanion/e_07.html, https://www.r-bloggers.com/2020/05/binary-logistic-regression-with-r/, assumptions from http://www.biostathandbook.com/multiplelogistic.html.

##Prepare variables for model

Make new variable for length of hibernation and code as numeric. 

```{r}
matingdata <- matingdata %>% 
  mutate(hib_length = case_when(treatment == "4 month" ~ '3',
                                treatment == "Recovery C" ~ '3',
                                treatment == "Recovery G" ~ '3',
                                treatment == "1 month" ~ '1',
                                treatment == "2 month" ~ '2'))

matingdata$hib_length <- as.numeric(matingdata$hib_length)

str(matingdata)
```

Select desired variables and save to dataframe.

```{r}
matingdata_survival_numeric <- matingdata %>% 
  select(day_0_weight, day_10_weight, mating_type, prehib_weightdiff, rel_prehib_weightdiff, status, hib_length)

str(matingdata_survival_numeric)
```

Make all non-numeric variables numeric.

```{r}
matingdata_survival_numeric$mating_type_num <- as.numeric(matingdata_survival_numeric$mating_type)
#1 = mated, 2 = unmated

matingdata_survival_numeric$status_num <- as.numeric(matingdata_survival_numeric$status)
#1 = alive, 2 = dead

#change the 2 for dead to a zero
matingdata_survival_numeric$status_num <-
  replace(matingdata_survival_numeric$status_num,matingdata_survival_numeric$status_num == 2,0)

#use str to double check that variables are coded properly
str(matingdata_survival_numeric)
```

##Correlation analysis

Remove factors from the data set for the correlation analysis.

```{r}
matingdata_survival_numeric_nofac <- matingdata_survival_numeric[-c(3,6)]
str(matingdata_survival_numeric_nofac)
```


Examine correlation among variables.

```{r}
chart.Correlation(matingdata_survival_numeric_nofac, method = "spearman", histogram = TRUE)
```


##Define model and model selection

```{r}
matingmodel3 <- glm(status_num ~ mating_type_num + day_0_weight + hib_length + rel_prehib_weightdiff, data = matingdata_survival_numeric_nofac, family= binomial)
```

###Variance inflation factors to detect multi-colinearity.

```{r}
#from car package
vif(matingmodel3)
```

No evidence of colinearity

###Plot residuals

```{r}
plot(fitted(matingmodel3),
     rstandard(matingmodel3))
```

###Check for overdispersion

Overdispersion is a situation where the residual deviance of the glm is large relative to the residual degrees of freedom. One guideline is that if the ratio of the residual deviance to the residual degrees of freedom exceeds 1.5, then the model is overdispersed. 

```{r}
summary(matingmodel3)$deviance / summary(matingmodel3)$df.residual
```

##Results of Model 3

```{r}
summary(matingmodel3)
```

